\section{The Approach to Calculus to be used in this book}\label{sec:1.1.6}

\begin{note}
  A thorough and complete treatment of either integral or differential calculus depends ultimately on a careful study of the real number system.
  This study in itself, when carried out in full, is an interesting but somewhat lengthy program that requires a small volume for its complete exposition.
  The approach in this book is to begin with the real numbers as undefined abjects and simply to list a number of fundamental properties of real numbers which we shall take as axioms.
  These axioms and some of the simplest theorems that can be deduced from them are discussed in \cref{ch:1.3}.

  Most of the properties of real numbers discussed here are probably familiar to the reader from his study of elementary algebra.
  However, there are a few properties of real numbers that do not ordinarily corne into consideration in elementary algebra but which play an important role in the calculus.
  These properties stem from the so-called \textbf{least-upper-bound axiom} (also known as the \textbf{completeness} or \textbf{continuity axiom}).

  To develop calculus as a complete, formal mathematical theory, it would be necessary to state, in addition to the axioms for the real number system, a list of the various ``methods of proof'' which would be permitted for the purpose of deducing theorems from the axioms.
  Every statement in the theory would then have to be justified either as an ``established law''
  (that is, an axiom, a definition, or a previously proved theorem) or as the result of applying one of the acceptable methods of proof to an established law.
  A program of this sort would be extremely long and tedious and would add very little to a beginner's understanding of the subject.
  Fortunately, it is not necessary to proceed in this fashion in order to get a good understanding and a good working knowledge of calculus.
  In this book the subject is introduced in an informal way, and ample use is made of geometric intuition whenever it is convenient to do so.
  At the same time, the discussion proceeds in a manner that is consistent with modern standards of precision and clarity of thought.
  All the important theorems of the subject are explicitly stated and rigorously proved.
\end{note}
